\appendices
%
% If you only have one appendix, you should change the above to:
%\appendix
%

\chapter{MORE INFORMATION ON EQUATIONS}

To demonstrate how an appendix should be inserted into the thesis
we have provided two appendices. This first appendix illustrates some
more advanced techniques to improve the appearance of your equations.
Below is a system of partial differential equations from a model for
cellular control by an external nutrient. The equations are complicated
and \LaTeX\ tends to allow them to run into each other. To prevent this
additional spacing between lines is achieved by using 
{\tt $\backslash\backslash$[3pt]} at the end of a line.  
Not also the use of the {\tt $\backslash$dfrac} command for fractions
in displayed equations.
The system is numbered only once by (\ref{sys2}), and
this is centered as best we can on one line.  The other lines have a
{\tt $\backslash$notag} command to suppress numbering.
\begin{align}
  \dot{U}_1(t) & = \tilde f(W_1(t-T)) - U_1(t) + \gamma_1U_2(R\sigma,
   t),	\notag \\[3pt]
  \dot{W}_1(t) & =  -\hat b_3W_1(t) + \gamma_3W_2(R\sigma,
   t),\notag \\[3pt]
  \frac{\partial U_2}{\partial t} & =  D_1\nabla^2U_2 - U_2 - \tilde f(W_1
    (t-T)) - \gamma_1U_2(R\sigma,t){\vrule width 0in depth .1in},
	\label{sys2} \\[3pt]
  \frac{\partial V_2}{\partial t} & =  D_2\nabla^2V_2 - b_2V_2 + c_0
    \bigl(U_2 + U_1(t)\bigr), \\[3pt]
  \frac{\partial W_2}{\partial t} & =  D_3\nabla^2W_2 - b_3W_2 + (\hat b_3
    -b_3)W_1 - \gamma_3W_2(R\sigma,t) \\
    &  \quad + k\left[\left[{\left(\frac{D_3}{r^2}\right)}\frac{d}{dr}\left(r^2
	   \frac{dh}{dr}\right) - b_3h\right]V_2(R,t) - h\dot V_2(R,t)
	   \right], 
\end{align}
For $t > 0$ and $R\sigma < r < R$ and with the boundary conditions:
\begin{align*}
 \dfrac{\partial U_2(R\sigma,t)}{\partial r} & = 
 \beta_1U_2(R\sigma,t), & 
 \dfrac{\partial U_2(R,t)}{\partial r} & =  0, \\
 \dfrac{\partial V_2(R\sigma,t)}{\partial r} & =  0, &
 \dfrac{\partial V_2(R,t)}{\partial r} & =  0,  \\
 \dfrac{\partial W_2(R\sigma,t)}{\partial r} & = 
   \beta_3W_2(R\sigma,t), & 
 \dfrac{\partial W_2(R,t)}{\partial r} & =    0.
\end{align*}

In algebra, one often uses commutative diagrams.  The packages
{\tt amscd}, {\tt diagram} and {\tt xypic} allow you to create
something like the following diagram below.
The {\tt xypic} package is by far the most powerful.
\begin{equation}
\xymatrix{
 X \times_Z Y \ar[d]^q \ar[r]_p & X \ar[d]_f       \\ 
 Y \ar[r]^g   & Z  
}
\end{equation}


The appendix can also include technical theorems and lemmas which are
call in the same manner as before. For example,
\begin{theorem}
The system of equations {\rm (\ref{sys2})} can exhibit periodic
solutions for certain parameter values.
\end{theorem}

\begin{proof}
The argument uses Hopf bifurcation techniques and is very
complicated. See Mahaffy {\it et al} \cite{MJV}.
\end{proof}

\chapter{LISTS AND QUOTATIONS}

To create a list you will want to use one of {\it itemize, enumerate, or
description}. For example:
\begin{description}
\item[continuous] A function $f$ is {\bf continuous} at $x$ if and only
if for every $\varepsilon >0$ there exists a $\delta(x) >0$ such that
whenever $|y-x|<\delta$, $|f(y)-f(x)| < \varepsilon$.
\item[uniformly continuous] A function $f$ is {\bf uniformly
continuous} if and only if for every $\varepsilon >0$ there exists a
$\delta >0$ such that whenever $|y-x|<\delta$, $|f(y)-f(x)| <
\varepsilon$ independent of $x$ and $y$.
\item[equicontinuous] A family of functions $f_n$ is {\bf
equicontinuous} at a point $x$ if and only if for every $\varepsilon >0$
there exists a $\delta >0$ such that whenever $|y-x|<\delta$,
$|f_n(y)-f_n(x)| < \varepsilon$ for all functions $f_n$.
\end{description}

\LaTeX\ provides an environment for block quotations. To agree with the
thesis manual follow the format below for a quotation exceeding four
lines. From Lewis Carrol's {\it Hunting of the Snark} we hear the
Bellman tell his crew:
 \vspace{.12pt}

{
\ssp
\begin{verse}
The Bellman himself they all praised to the skies--\\
Such a carriage, such ease and such grace!\\
Such solemnity, too! One could see he was wise,\\
The moment one looked in his face!\\
 \vspace{.15in}
He had bought a large map representing the sea,\\
Without the least vestige of land:\\
And the crew were much pleased when they found it to be\\
A map they could all understand.\\
 \vspace{.15in}
``What's the good of Mercator's, North Poles and Equators,\\
Tropics, Zones, and Meridian Lines?''\\
So the Bellman would cry: and the crew would reply,\\
``They are merely conventional signs!''\\
 \vspace{.15in}
``Other maps are such shapes, with their islands and capes!\\
But we've got our brave Captain to thank''\\
(So the crew would protest) ``that he's bought us the best--\\
A perfect and absolute blank!''\\
\end{verse}
}

